Type-Safe Modular Parsing

### 2017-04-24 SLE_17/sections/CaseStudy.tex
Code & content

\begin{figure}[ht]
\lstinputlisting[linerange=15-33]{../Scala/Parser/src/PaperCode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_VARAPP
\caption{The \inlinecode{VarApp} component in case study.}\label{fig:casestudy-varapp}
\end{figure}

@@

\lstinputlisting[linerange=9-11]{../Scala/Parser/src/PaperCode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_EPARSER

@@

\begin{figure}[t]
\lstinputlisting[linerange=37-64]{../Scala/Parser/src/PaperCode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_UNTYPED
\caption{Build the \inlinecode{Untyped} calculus by composing to language components.}\label{fig:casestudy-untyped}
\end{figure}

@@

\lstinputlisting[linerange=68-78]{../Scala/Parser/src/PaperCode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_DEMO

>>>>

\begin{figure}[ht]
\lstinputlisting[linerange=15-33]{code/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_VARAPP
\caption{The \inlinecode{VarApp} component in case study.}\label{fig:casestudy-varapp}
\end{figure}

@@

\lstinputlisting[linerange=9-11]{code/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_EPARSER

@@

\begin{figure}[t]
\lstinputlisting[linerange=37-64]{code/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_UNTYPED
\caption{Build the \inlinecode{Untyped} calculus by composing to language components.}\label{fig:casestudy-untyped}
\end{figure}

@@

\lstinputlisting[linerange=68-78]{code/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_DEMO

### 2017-04-24 SLE_17/sections/Discussion.tex
Code & content

Using Object Algebras, it is easy to
model multi-sorted languages. If the syntax contains multiple sorts, we can distinguish them by different type parameters. As an example, we extend the language of literals, additions and variables by introducing types, including primitive type \inlinecode{int} and function (arrow) type, and we add lambda abstractions to expressions. Now the language has two sorts in the syntax, types and expressions:\\

\begin{tabular}{m{0.45\linewidth}m{0.45\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<type> ::= `int' \alt <type> `->' <type>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::=  ... \alt `\\' <ident> `:' <type> `.' <expr>
\end{grammar}
\end{tabular}

@@

\begin{figure}[ht]
\lstinputlisting[linerange=31-59]{../Scala/Parser/src/PaperCode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
\caption{Code for parsing a language with multi-sorted syntax for types and expressions.}
\label{fig:multi}
\end{figure}

@@

\lstinputlisting[linerange=63-66]{../Scala/Parser/src/PaperCode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX_CLIENT
The code works as expected to parse and pretty-print the result.

@@

\lstinputlisting[linerange=76-81]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=BASEPARSER_UNTYPEDLAM

@@

\lstinputlisting[linerange=85-90]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=EXTPARSER_TYPEDLAM

@@

\begin{comment}
It is natural to keep overriding existing parsers, whereas another potential use of this pattern is to reuse old versions of a parser. For instance,
we have a \lstinline{NewParser} that extends \lstinline{ExtParser}, with some more cases added, but for the lambda case we want to go back to the untyped one.
It can be achieved by instantiating an instance of \lstinline{BaseParser} to obtain the old \lstinline{pLam}. However, we must have \inlinecode{lazy} modifier for all the \inlinecode{pLam} starting from \inlinecode{BaseParser}.

\lstinputlisting[linerange=94-102]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=NEWPARSER_UNTYPEDLAM
\end{comment}

@@

\begin{itemize}
    \item \textbf{Object Algebra interface:} defined as a trait for the abstract syntax. The type parameters represent multiple sorts of syntax, and the methods are constructs.
    \item \textbf{Parser:} corresponding parser of the abstract syntax, written in a modular way as we demonstrated in previous sections.
    \item \textbf{Object Algebras (optional):} concrete operations on ASTs, such as pretty-printing.
\end{itemize}

@@

\begin{figure}[t]
\begin{tabular}{m{0.42\linewidth}m{0.52\linewidth}}
\lstinputlisting[linerange=113-132]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VAREXPR
&
\lstinputlisting[linerange=136-156]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_TYPEDLAM
\end{tabular}
\caption{Language components \lstinline{VarExpr} (left) and \lstinline{TypedLam} (right).}\label{fig:lng-components}
\end{figure}

@@

\begin{figure}[t]
\lstinputlisting[linerange=160-170]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
\caption{Reuse two language components \lstinline{VarExpr} and \lstinline{TypedLam}.}\label{fig:compose-components}
\end{figure}

@@

\begin{itemize}

\item {\bf Parsing Technique}

Our approach does not depend on a particular parser combinator library
or parsing algorithm. We demonstrated that Packrat parsing has some
advantages and is suitable in a modular setting. However, under
certain circumstances or in other programming languages, it may not be
the best choice. One can use other parser combinator libraries as long
as the guidelines proposed in Section~\ref{sec:packrat} are met.

\item {\bf Composing and Extending Parsers}

  We use traits to model parsers, OO inheritance to compose them, and
  method overriding for their extensibility. These language features
  may not be supported in other programming languages, especially in
  functional languages such as Haskell. Nevertheless, \textit{open
    recursion}~\cite{CookThesis} could be used as an alternative. An explicit
  ``self-reference'' parameter is able to explain the recursive calls
  dynamically, by the real argument passed at runtime.

\item {\bf Extensible ASTs}

Besides Object Algebras, many other techniques including \textit{Data
  types à la carte} (DTC)~\cite{swierstra2008data} and the Cake pattern~\cite{odersky2005independently} also support extensible data structures. These alternatives could also be adopted to build ASTs and corresponding parsers.

@@

As we demonstrated, the whole modular parsing framework itself is
modular, which means it can be customized easily.
%We conducted some previous experiments of modular parsing in Haskell, based on open recursion, MRM and Parsec. However, we suffered from issues of Parsec, as illustrated in Section~\ref{subsec:challenges}.

>>>>

Using Object Algebras, it is easy to
model multi-sorted languages. If the syntax contains multiple sorts, we can distinguish them by different type parameters. As an example, we extend the language of literals, additions and variables by introducing types, including primitive type \inlinecode{int} and function (arrow) type, and we add lambda abstractions to expressions. Now the language has two sorts in the syntax, types and expressions:

\setlength{\grammarindent}{5em}
\begin{grammar}
<type> ::= `int' \alt <type> `->' <type>

<expr> ::=  ... \alt `\\' <ident> `:' <type> `.' <expr>
\end{grammar}

@@

\begin{figure}[ht]
\lstinputlisting[linerange=31-59]{code/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
\caption{Code for parsing a language with multi-sorted syntax for types and expressions.}
\label{fig:multi}
\end{figure}

@@

\lstinputlisting[linerange=63-66]{code/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX_CLIENT
The code works as expected to parse and pretty-print the result.

@@

\lstinputlisting[linerange=76-81]{code/Sec5/Code1.scala}% APPLY:linerange=BASEPARSER_UNTYPEDLAM

@@

\lstinputlisting[linerange=85-90]{code/Sec5/Code1.scala}% APPLY:linerange=EXTPARSER_TYPEDLAM

@@

@@

\begin{itemize}[leftmargin=*]
    \item \textbf{Object Algebra interface:} defined as a trait for the abstract syntax. The type parameters represent multiple sorts of syntax, and the methods are constructs.
    \item \textbf{Parser:} corresponding parser of the abstract syntax, written in a modular way as we demonstrated in previous sections.
    \item \textbf{Object Algebras (optional):} concrete operations on ASTs, such as pretty-printing.
\end{itemize}

@@

\begin{figure}[t]
\lstinputlisting[linerange=113-132]{code/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VAREXPR
\lstinputlisting[linerange=136-156]{code/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_TYPEDLAM
\caption{Language components \lstinline{VarExpr} (\huang{todo}left) and \lstinline{TypedLam} (right).}\label{fig:lng-components}
\end{figure}

@@

\begin{figure}[t]
\lstinputlisting[linerange=160-170]{code/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
\caption{Reuse two language components \lstinline{VarExpr} and \lstinline{TypedLam}.}\label{fig:compose-components}
\end{figure}

@@

\begin{itemize}[leftmargin=*]

\item {\bf Parsing Technique:}
Our approach does not depend on a particular parser combinator library
or parsing algorithm. We demonstrated that Packrat parsing has some
advantages and is suitable in a modular setting. However, under
certain circumstances or in other programming languages, it may not be
the best choice. One can use other parser combinator libraries as long
as the guidelines proposed in Section~\ref{sec:packrat} are met.

\item {\bf Composing and Extending Parsers:}
  We use traits to model parsers, OO inheritance to compose them, and
  method overriding for their extensibility. These language features
  may not be supported in other programming languages, especially in
  functional languages such as Haskell. Nevertheless, \textit{open
    recursion}~\cite{CookThesis} could be used as an alternative. An explicit
  ``self-reference'' parameter is able to explain the recursive calls
  dynamically, by the real argument passed at runtime.

\item {\bf Extensible ASTs:}
Besides Object Algebras, many other techniques including \textit{Data
  types à la carte} (DTC)~\cite{swierstra2008data} and the Cake pattern~\cite{odersky2005independently} also support extensible data structures. These alternatives could also be adopted to build ASTs and corresponding parsers.

@@

As we demonstrated, the whole modular parsing framework itself is
modular, which means it can be customized easily.

### 2017-04-24 SLE_17/sections/Inheritance.tex
Code & content

In the remainder of this section we show a technique for
defining \emph{extensible} parsing code. That is new
language constructs can be modularly added without changing
existing parsing code. The technique employs standard OO
style mechanisms such as subtyping and inheritance, as well as,
Packrat parsing combinators, and enables a better integration between
parsing and OO ASTs.

@@

\lstinputlisting[linerange=29-31]{../Scala/Parser/src/PaperCode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_SIMPLE_LAM

@@

\lstinputlisting[linerange=35-38]{../Scala/Parser/src/PaperCode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_BAD_ATTEMPT

@@

\lstinputlisting[linerange=42-47]{../Scala/Parser/src/PaperCode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_APPROACH

@@

\lstinputlisting[linerange=67-73]{../Scala/Parser/src/PaperCode/Sec3Inheritance/Code1.scala}% APPLY:linerange=MULTIPLE_INHERITANCE

@@

\lstinputlisting[linerange=79-95]{../Scala/Parser/src/PaperCode/Sec3Inheritance/Code1.scala}% APPLY:linerange=BOOL_EXAMPLE

>>>>

In the remainder of this section we show a technique for
defining \emph{extensible} parsing code. That is new
language constructs can be modularly added without changing
existing parsing code. The technique employs standard OO
style mechanisms such as subtyping and inheritance, as well as,
Packrat parser combinators, and enables a better integration between
parsing and OO ASTs.

@@

\lstinputlisting[linerange=29-31]{code/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_SIMPLE_LAM

@@

\lstinputlisting[linerange=35-38]{code/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_BAD_ATTEMPT

@@

\lstinputlisting[linerange=42-47]{code/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_APPROACH

@@

\lstinputlisting[linerange=67-73]{code/Sec3Inheritance/Code1.scala}% APPLY:linerange=MULTIPLE_INHERITANCE

@@

\lstinputlisting[linerange=79-95]{code/Sec3Inheritance/Code1.scala}% APPLY:linerange=BOOL_EXAMPLE

### 2017-04-24 SLE_17/sections/ObjectAlgebras.tex
Code & content

\lstinputlisting[linerange=23-23]{../Scala/Parser/src/PaperCode/Sec4OA/Code1.scala}% APPLY:linerange=FVARSEXPR

@@

\lstinputlisting[linerange=27-37]{../Scala/Parser/src/PaperCode/Sec4OA/Code1.scala}% APPLY:linerange=FVAR_CLASSES

@@

\lstinputlisting[linerange=5-8]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALG
We have two constructors \inlinecode{lit} and \inlinecode{add}, corresponding to literals and additions. The trait \inlinecode{ExprAlg} is called an \textit{Object Algebra interface}. The trait is parametrized by the type
\inlinecode{E}, which abstracts over the concrete type of the AST.

@@

\lstinputlisting[linerange=12-15]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
Here \inlinecode{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
One can define evaluation operation as a new trait on this expression language:

\lstinputlisting[linerange=19-22]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EVAL

@@

\lstinputlisting[linerange=26-28]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALGEXT
Now pretty-printing on the new language can be realized without modifying existing code:

\lstinputlisting[linerange=32-34]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
An observation is that only the new case is implemented for pretty-printing, and the others have been inherited.
In other words, existing code was reused and was not modified!

@@

\lstinputlisting[linerange=41-41]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_MAKEEXP

@@

\lstinputlisting[linerange=8-16]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser works fine, but it is not extensible! For example, we have demonstrated in Section~\ref{subsec:overriding} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method overriding for \inlinecode{pExpr} would require a type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]}, where the type of the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This difference prohibits method/field overriding. Scala supports type-refinement of fields if the overriding
field has a type which is a \emph{subtype} of the original field.
Unfortunately, in this case, although \inlinecode{VarExprAlg} is a subtype
of \inlinecode{ExprAlg}, the function type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]} is a supertype of \inlinecode{ExprAlg[E] =>} \inlinecode{Parser[E]}. Thus it is not possible to have overriding.

@@

\lstinputlisting[linerange=23-30]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER
This is precisely the pattern that we advocate for modular parsing.
One important remark is we introduce \inlinecode{pE} for recursive calls.
The reason why we use an extra and seemingly redundant field, is due to a subtle issue caused by Scala language and its parser combinator library. There is a restriction of \inlinecode{super} keyword in Scala that \inlinecode{super} can only use methods defined by keyword \inlinecode{def}, but cannot access fields defined by \inlinecode{val}, while the parser combinator library suggests using \inlinecode{val} to define parsers, especially for left-recursive ones. Our workaround is that we use different synonyms for \inlinecode{pE} in different traits, so that we can directly distinguish them by names without using \inlinecode{super}.

@@

\lstinputlisting[linerange=34-39]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER

@@

\lstinputlisting[linerange=43-46]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER_CLIENT

@@

\lstinputlisting[linerange=45-56]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_REFACTOR
Here the \inlinecode{VarExprAlg} expression \inlinecode{1 + x} is changed to \inlinecode{ExprAlg} expression \inlinecode{1 + 2}, before applying the pretty-printing algebra. The abstract field
\inlinecode{alg} in \inlinecode{Refactor} could be any algebra simply for delegation. With an algebra instance of \inlinecode{alg}, \inlinecode{Refactor} is able to transform a \inlinecode{VarExprAlg} expression before producing the result. Furthermore, transformations can be composed in a linear pipeline-style beforehand~\cite{Zhang2015}.

>>>>

\lstinputlisting[linerange=23-23]{code/Sec4OA/Code1.scala}% APPLY:linerange=FVARSEXPR

@@

\lstinputlisting[linerange=27-37]{code/Sec4OA/Code1.scala}% APPLY:linerange=FVAR_CLASSES

@@

\lstinputlisting[linerange=5-8]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALG
We have two constructors \inlinecode{lit} and \inlinecode{add}, corresponding to literals and additions. The trait \inlinecode{ExprAlg} is called an \textit{Object Algebra interface}. The trait is parametrized by the type
\inlinecode{E}, which abstracts over the concrete type of the AST.

@@

\lstinputlisting[linerange=12-15]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
Here \inlinecode{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
One can define evaluation operation as a new trait on this expression language:

\lstinputlisting[linerange=19-22]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EVAL

@@

\lstinputlisting[linerange=26-28]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALGEXT
Now pretty-printing on the new language can be realized without modifying existing code:

\lstinputlisting[linerange=32-34]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
An observation is that only the new case is implemented for pretty-printing, and the others have been inherited.
In other words, existing code was reused and was not modified!

@@

\lstinputlisting[linerange=41-41]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_MAKEEXP

@@

\lstinputlisting[linerange=8-16]{code/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser works fine, but it is not extensible! For example, we have demonstrated in Section~\ref{subsec:overriding} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method overriding for \inlinecode{pExpr} would require a type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]}, where the type of the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This difference prohibits method/field overriding. Scala supports type-refinement of fields if the overriding
field has a type which is a \emph{subtype} of the original field.
Unfortunately, in this case, although \inlinecode{VarExprAlg} is a subtype
of \inlinecode{ExprAlg}, the function type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]} is a supertype of \inlinecode{ExprAlg[E] =>} \inlinecode{Parser[E]}. Thus it is not possible to have overriding.

@@

\lstinputlisting[linerange=23-30]{code/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER
This is precisely the pattern that we advocate for modular parsing.
One important remark is we introduce \inlinecode{pE} for recursive calls.
The reason why we use an extra and seemingly redundant field, is due to a subtle issue caused by Scala language and its parser combinator library. There is a restriction of \inlinecode{super} keyword in Scala that \inlinecode{super} can only use methods defined by keyword \inlinecode{def}, but cannot access fields defined by \inlinecode{val}, while the parser combinator library suggests using \inlinecode{val} to define parsers, especially for left-recursive ones. Our workaround is that we use different synonyms for \inlinecode{pE} in different traits, so that we can directly distinguish them by names without using \inlinecode{super}.

@@

\lstinputlisting[linerange=34-39]{code/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER

@@

\lstinputlisting[linerange=43-46]{code/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER_CLIENT

@@

\lstinputlisting[linerange=45-56]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_REFACTOR
Here the \inlinecode{VarExprAlg} expression \inlinecode{1 + x} is changed to \inlinecode{ExprAlg} expression \inlinecode{1 + 2}, before applying the pretty-printing algebra. The abstract field
\inlinecode{alg} in \inlinecode{Refactor} could be any algebra simply for delegation. With an algebra instance of \inlinecode{alg}, \inlinecode{Refactor} is able to transform a \inlinecode{VarExprAlg} expression before producing the result. Furthermore, transformations can be composed in a linear pipeline-style beforehand~\cite{Zhang2015}.

### 2017-04-24 SLE_17/sections/PackratParsing.tex
Code & content

\begin{itemize}

\item {\bf Modular parsers should be left-recursive;}

\item {\bf Modular parsers should use a longest match composition operator.}

@@

\begin{tabular}{m{0.4\linewidth}m{0.5\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int>
\end{grammar}
&
\begin{lstlisting}[language=PlainCode]
parseExpr = parseInt
\end{lstlisting}
\end{tabular}

@@

\begin{tabular}{m{0.4\linewidth}m{0.5\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <expr> `-' <int>
\end{grammar}
&
\begin{lstlisting}[language=PlainCode]
parseExpr = parseSub <|> parseInt
parseSub = do
  e <- parseExpr
  ...
\end{lstlisting}
\end{tabular}

To solve this issue, we have to rewrite the grammar as below to eliminate left-recursion and build the parser based on this new grammar.\\

\begin{tabular}{m{0.4\linewidth}m{0.5\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> <expr'>

<expr'> ::= <empty> \alt `-' <int> <expr'>
\end{grammar}
&
\end{tabular}

@@

\begin{figure}[t]
\centering
\lstinputlisting[linerange=42-53]{../Scala/Parser/src/PaperCode/Sec2Packrat/Code2.scala}% APPLY:linerange=PACKRAT_PAPERCODE
\caption{Helper object for code demonstration in this paper.}\label{fig:papercode}
\end{figure}

@@

\lstinputlisting[linerange=9-17]{../Scala/Parser/src/PaperCode/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_OO_AST

@@

\lstinputlisting[linerange=21-29]{../Scala/Parser/src/PaperCode/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_SIMPLE_EXPR

@@

\begin{table}[t]
\begin{tabular}{l}
\hline
\begin{lstlisting}
def ~[U](q: => Parser[U]): Parser[~[T, U]]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition. \\
\hline
\begin{lstlisting}
def ^^[U](f: (T) => U): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for function application. \\
\hline
\begin{lstlisting}
def ^^^[U](v: => U): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator that changes a successful result into the specified value. \\
\hline
\begin{lstlisting}
def <~[U](q: => Parser[U]): Parser[T]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition which keeps only the left result. \\
\hline
\begin{lstlisting}
def ~>[U](q: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition which keeps only the right result. \\
\hline
\begin{lstlisting}
def repsep[T](p: => Parser[T], q: => Parser[Any]): Parser[List[T]]
\end{lstlisting} \\
\hspace{.2in}- A parser generator for interleaved repetitions. \\
\hline
\begin{lstlisting}
def ident: Parser[String]
\end{lstlisting} \\
\hspace{.2in}- A parser which matches an identifier. \\
\hline
\begin{lstlisting}
def numericLit: Parser[String]
\end{lstlisting} \\
\hspace{.2in}- A parser which matches a numeric literal. \\
\hline
\begin{lstlisting}
def |[U >: T](q: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for alternative composition. \\
\hline
\begin{lstlisting}
def |||[U >: T](q0: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for alternative with longest match composition. \\
\hline \\
\end{tabular}
\caption{Common combinators from the Scala standard parser combinator library.}\label{tab:packrat}
\end{table}

@@

\lstinputlisting[linerange=33-33]{../Scala/Parser/src/PaperCode/Sec2Packrat/Code2.scala}% APPLY:linerange=PACKRAT_RUNPARSER

>>>>

\begin{itemize}[leftmargin=*]

\item \textbf{Modular parsers should support left-recursion.}

\item \textbf{Modular parsers should use a longest match composition operator.}

@@

\begin{minipage}{10em}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int>
\end{grammar}
\end{minipage}
\begin{minipage}{10em}
\begin{lstlisting}[language=PlainCode]
parseExpr = parseInt
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{12em}
\setlength{\grammarindent}{4em}
\begin{grammar}
<expr> ::= <int> \alt <expr> `-' <int>
\end{grammar}
\end{minipage}
\begin{minipage}{15em}
\begin{lstlisting}[language=PlainCode]
parseExpr =
  parseSub <|> parseInt

parseSub = do
  e <- parseExpr
  ...
\end{lstlisting}
\end{minipage}

To solve this issue, we have to rewrite the grammar as below to eliminate left-recursion and build the parser based on this new grammar.

\vspace{0.4\baselineskip}
\begin{minipage}{15em}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> <expr'>

<expr'> ::= <empty> \alt `-' <int> <expr'>
\end{grammar}
\end{minipage}
\vspace{0.4\baselineskip}

@@

\begin{figure}[t]
\centering
\lstinputlisting[linerange=42-53]{code/Sec2Packrat/Code2.scala}% APPLY:linerange=PACKRAT_PAPERCODE
\caption{Helper object for code demonstration in this paper.}\label{fig:papercode}
\end{figure}

@@

\lstinputlisting[linerange=9-17]{code/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_OO_AST

@@

\lstinputlisting[linerange=21-29]{code/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_SIMPLE_EXPR

@@

\begin{table}
\begin{tabularx}{\linewidth}{X}
\hline
\lstinline{def ~[U](q: => Parser[U]): Parser[~[T, U]]} \\
- A parser combinator for sequential composition. \\ \hline
\lstinline{def \^\^[U](f: (T) => U): Parser[U]} \\
- A parser combinator for function application. \\ \hline
\lstinline{def \^\^\^[U](v: => U): Parser[U]} \\
- A parser combinator that changes a successful result into the specified value. \\ \hline
\lstinline{def <~[U](q: => Parser[U]): Parser[T]} \\
- A parser combinator for sequential composition which keeps only the left result. \\ \hline
\lstinline{def ~>[U](q: => Parser[U]): Parser[U]} \\
- A parser combinator for sequential composition which keeps only the right result. \\ \hline
\lstinline{def ident: Parser[String]} \\
- A parser which matches an identifier. \\ \hline
\lstinline{def numericLit: Parser[String]} \\
- A parser which matches a numeric literal. \\ \hline
\lstinline{def |[U >: T](q: => Parser[U]): Parser[U]} \\
- A parser combinator for alternative composition. \\ \hline
\lstinline{def |||[U >: T](q0: => Parser[U]): Parser[U]} \\
- A parser combinator for alternative with longest match composition. \\ \hline
\end{tabularx}
\caption{Common combinators from the Scala standard parser combinator library.}\label{tab:packrat}
\end{table}

@@

\lstinputlisting[linerange=33-33]{code/Sec2Packrat/Code2.scala}% APPLY:linerange=PACKRAT_RUNPARSER
